---
title: "STAT508 Individual Midterm"
author: "Callum Arnold"
output:
    html_notebook:
        code_folding: hide
        toc: yes
        toc_float: yes
---

# Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(ggforce)
library(here)
library(caret)
library(MASS)
library(class)
library(kableExtra)
library(hrbrthemes)
library(janitor)
library(broom)
library(rsample)
library(corrplot)
library(lares)

RNGkind(sample.kind = "Rounding")
set.seed(1)

theme_set(theme_ipsum())
```

```{r}
wiki <- read.csv(
  here("data", "wiki4HE.csv"),
  header = TRUE,
  sep = ";",
  na.strings = "?"
)
```

# Data Cleaning and Exploration

```{r}
wiki <- wiki %>%
    clean_names()

names(wiki)

wiki <- wiki %>%
    rename(phd = ph_d)
```

```{r}
summary(wiki)
```

```{r}
hist_plts <- function(data, var, ncol, nrow, pages) {
  for (i in 1:pages) {
    plot <- data %>%
      ggplot(
        aes(
          x = {{ var }}
        )
      ) +
      geom_histogram(binwidth = 1) +
      labs(
        caption = paste0("Page ", i, " of ", pages)
      ) +
      scale_x_continuous() +
      scale_fill_ipsum() +
      facet_wrap_paginate(
        ~Variable,
        scales = "free",
        nrow = {{ ncol }},
        ncol = {{ nrow }},
        page = i) +
      theme_ipsum_rc()

    print(plot)
  }
}
```

```{r, fig.width=14}
wiki %>%
    pivot_longer(
        cols = everything(),
        names_to = "Variable",
        values_to = "Score"
        ) %>%
    hist_plts(var = Score, ncol = 3, nrow = 3, pages = 6)
```

```{r}
wiki_trans <- wiki %>%
    mutate(yearsexp = log1p(yearsexp))

ggplot(wiki_trans) +
    geom_histogram(aes(x = yearsexp), binwidth = 0.5)
```


```{r, fig.width=10}
corr_matrix <- cor(wiki_trans[, -c(6, 7, 8)], use = "complete.obs")
corrplot(corr_matrix, order = "hclust", type = "upper")
```

```{r}
corr_cross(df = wiki_trans, top = 20)
```


# Part 1
## PCA

```{r}
wiki_pca <- wiki_trans %>%
  dplyr::select(!c(use1:use5, age, phd, university)) %>%
  na.omit() %>%
  prcomp(center=TRUE, scale = TRUE)

summary(wiki_pca)
```
```{r}
variance <- (wiki_pca$sdev)^2

pve <- variance / sum(variance)
```

```{r}
par(mfrow = c(1, 2))

barplot(pve, xlab = "PC", ylab = "Proportion of variance explained", names.arg = 1:length(pve), las = 1, col = "gray")
barplot(cumsum(pve), xlab = "Principal Component", ylab = " Cumulative Proportion of Variance Explained", ylim = c(0, 1))
```

Elbow around PC6.

# Part 2
## Data Transformation

Transform Likert responses "1-3" as "0" (No), and "4-5" as "1" (Yes). A response of "3" is transformed to "No" as each of the survey items asks a positive actions that indicates an improvement over the status quo, therefore can be viewed as the alternative hypothesis, and an indifferent response is like failing to reject the null.

```{r}

```


## Logistic Regression
## LDA
## QDA
## kNN
## Model Comparison